From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/reversible1.tex
Progress
Formatting
Cite pi-dual

To see how this language expresses reversible circuits, we present a few examples using the Agda embedding of the
language~\cite{choudhuryArtifactSymmetriesReversible2021}. First it is possible to directly mimic the
\verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We use the type
$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define
using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus
the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

\cif{}

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the left branch, which corresponds to
the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
type~\ensuremath{A}. The right branch, which corresponds to the boolean being \textsf{false}, passes the value of type $A$
through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the previous section
is transcribed as follows:

\adder{}

@@

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second
circuit, instead of transcribing it directly, we express it using $\Afun{cif}$.

\resettwo{}

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

\leveltwoblockone{}

>>>>

To see how this language expresses reversible circuits, we present a few examples using the Agda embedding of the
language~\cite{choudhuryArtifactSymmetriesReversible2021,sabryJacquesCarettePidualSecond2021}. First it is possible to
directly mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We
use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define
using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus
the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:
\medskip
\cif{}

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the left branch, which corresponds to
the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
type~\ensuremath{A}. The right branch, which corresponds to the boolean being \textsf{false}, passes the value of type $A$
through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the previous section
is transcribed as follows:
\medskip
\adder{}

@@

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second
circuit, instead of transcribing it directly, we express it using $\Afun{cif}$.
\medskip
\resettwo{}

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:
\medskip
\leveltwoblockone{}
